---
title: "Statistical Inference - Assignment part 1"
author: "Christopher Lovell"
date: "Wednesday, January 21, 2015"
output: html_document
---

# Statistical Inference Course Project - Part 1
In this project we re going to cimluate the exponential distribution, and use the results to compare the properties of the sample and underlying distributions.


## Set Up
First we need to set up our environment. We load the ggplot2 library for doing our visualisations, and initialise our random seed so that our results are consistent on repeat.
```{r}
if(!require(ggplot2)){
  install.packages("ggplot2")
}

set.seed(1)
```

## Simulations

First we initialise lambda and our simulation values, where `N` is the total number of simulations, and `n` is the sample size.
```{r} 
lambda=0.2
N=1000
n=40
```

We can then create an `N*n` matrix of our simulation results.
```{r}
data<-matrix(rexp(N*n,lambda), N)
```

## Sample Mean versus Theoretical Mean

The theoretical mean is given by $\mu=1/\lambda$. 
```{r}
mean = 1/lambda
```

We can compare this to the sample mean by plotting the distribution of sample means for each simulation, and compare to the theoretical mean shown in red.
```{r}
sample_means<-apply(data,1,function(x) mean(x))
sample_mean<-mean(sample_means)

hist(sample_means,xlab = "Sample mean distribution",breaks=seq(3,9,0.2))
abline(v=mean,col="red")
```

The centre of the distribution alines with the theoretical mean.

*Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.*

## Sample Variance versus Theoretical Variance

The theoretical variance is given by $Var(X)=\frac{1}{\lambda\sqrt n}$. 
```{r}
sd = (1/lambda)/(n^0.5)
var=sd^2
```

We can compare this to the sample variance by computing the variance across the sample means.
```{r eval=FALSE}
var(sample_means)
``` 
This gives us a sample variance of `r var(sample_means)` which is similar to the theoretical variance of `r var`.

## Distribution
We can tell that the distribution is approximately normal by plotting a normal curve, with mean and standard deviation parameters taken from our theoretical model, over our sample distribution.

```{r}
sample_means<-as.data.frame(sample_means)
names(sample_means)<-"RowMeans"

# histogram of 1000 averages of 40 random variables 
cp<-ggplot(sample_means,aes(x=sample_means))
cp<-cp+geom_histogram(binwidth = lambda,fill="dark grey",color="white",aes(y = ..density..))
cp<-cp+stat_function(fun=dnorm,args=list(mean=mean,sd=sd))
cp<-cp+geom_vline(xintercept=mean,color="red")
cp<-cp+geom_vline(xintercept=sample_mean,color="blue")
cp
```

## Confidence Intervals
```{r}
#ci <- mean + (c(-1,1)*1.96*sd/n)
#sample_ci <- sample_mean + (c(-1,1)*1.96*sample_sd/n)


```


